---
title: "Quality Control for dplyr"
author: "Adam Maier, TNTP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Control for dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose and Overview

Most TNTP analysts use the package `dplyr` to clean data and perform basic analyses. And for good reason, it's a fantastic set of functions that makes this work substantially easier. However, to determine the extent to which the `dplyr` functions accomplished one's intended goals, analysts often have to write additional code to check their work, or *eyeball* their manipulated data to ensure correctness. For example, to see how many rows were lost after filtering data, one would have to remember the number of original rows in the data and then manually subtract the rows in the newly filtered data; to see how many rows were dropped for each group would take additional code. To see how many rows actually joined when performing a merge, or to see if the newly joined data contains more rows than expected because of duplicated matches, an analyst would have to write code before and after implementing a `dplyr` join function, like `left_join` or `full_join`.

The `reviewr` package makes performing these checks and others easy to implement and read. By simply adding the suffix `_qc` to the end of several core `dplyr` functions, you will get **the exact same result** as you would had you used the normal `dplyr` functions, but additional information about what just happened will automatically print. For example, using `filter_qc` instead of `filter` will filter the data exactly as `filter` will, but will also report how many (and what percent) of rows were dropped. This makes using `reviewr` easy and allows you to quickly toggle back and forth between regular `dplyr` functions and their `reviewr` counterparts. No wrappers around your code or any other set-up; just add `_qc`.

`reviewr`, therefore, allows you to get under the hood of data cleaning with `dplyr`. It does not attempt or intend to tell if you `dplyr` "worked", but rather whether how you used the `dplyr` function(s) matched what you intended. (Does it make sense that you dropped 20 rows when performing the filter? Did you expect the join to not be a perfect 1-1 join? etc.)

### Why use reviewr?

Some of the checks performed by `reviewr` are simple and are easy to do without the package. Determining how many rows were dropped after a filter is often just a matter of subtraction. Although reviewr saves you the step, it has the additional value of being more easily used in a series of piped functions. For example, the following chunk of code performs filters at two steps. To determine how many rows were dropped at each step would require the analyst to break apart the code and test both. But with `reviewr`, you can keep the pipe as is, and simply add `_qc`.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(reviewr)
```

```{r, echo = TRUE, warning = FALSE}

normal_output <- mtcars %>% 
  filter(mpg < 25) %>% 
  group_by(vs) %>% 
  filter(hp > mean(hp))
```

```{r, echo = TRUE}
reviewr_output <- mtcars %>% 
  filter_qc(mpg < 25) %>% 
  group_by(vs) %>% 
  filter_qc(hp > mean(hp))

```

The two outputs are identical; `reviewr` simply prints more information.

```{r, echo = TRUE}
all.equal(normal_output, reviewr_output)
```

Because `reviewr` was built on the premise of toggling back and forth between `dplyr` and `reviewr` functions when needed, it is meant to particularly assist those performing Quality Control checks, as the checker can simply add `_qc` at the end of the existing code to dive more deeply into what the code did.

### Working with grouped data

If you simply add `_qc_` to the end of `dplyr` code, `reviewr`could be helpful. But it also includes an option to learn more about what happened to each group in your data if you have a grouped structure. For example, how many rows in each group were dropped when filtering? Which groups, if any, had NA values created when doing a `group_by()` and `summarize()` combination, and on which variables? Many of the `reviewr` functions have an option to print this information:

```{r, echo = TRUE}
reviewr_output <- mtcars %>% 
  filter_qc(mpg < 25) %>% 
  group_by(vs) %>% 
  filter_qc(hp > mean(hp), .group_check = T)
```

Again, the two outputs are identical; `reviewr` simply prints more information.

```{r, echo = TRUE}
all.equal(normal_output, reviewr_output)
```

## reviewr Functions

There are currently four sets of `dplyr` functions to which you can add `_qc` suffix and get a `reviewr` equivalent

* Filter.  `filter_qc`, `filter_all_qc`, `filter_at_qc`, and `filter_if_qc`
* Mutate.  `mutate_qc`, `mutate_all_qc`, `mutate_at_qc`, and `mutate_if_qc`. And `transmute_qc`, `transmute_all_qc`, `transmute_at_qc`, and `transmute_if_qc`
* Summarize.  `summarize_qc`, `summarize_all_qc`, `summarize_at_qc`, and `summarize_if_qc`
* Joins.  `full_join_qc`, `inner_join_qc`, `left_join_qc`, `right_join_qc`, `semi_join_qc`, and `anti_join_qc`

Although, some `reviewr` functions have additional options, all `reviewr` functions will work by simply using them as you would `dplyr`. This means you can start using them now: if you know how to use their `dplyr` versions, the `reviewr` functions will be implemented identically.

Below are examples for each set of `reviewr` functions and a brief discussion of additional options you can use.

### Filtering

The set of `reviewr` filtering functions report the number of rows that were dropped after performing the filter.

```{r, echo = TRUE}
filter_qc(mtcars, gear == 4, mpg > 30)
```

If the data is grouped, you can set the option `.group_check = T` to see how many rows were dropped per group. Note that this could print a big table if you have a lot of groups. The default is `.group_check = F`

```{r, echo = TRUE}
result <- mtcars %>%
  group_by(vs) %>%
  filter_qc(gear == 4, mpg > 30, .group_check = T)
```

`reviewr` has  `_qc` versions of the scoped filter functions, too.

```{r, echo = TRUE}
filter_at_qc(mtcars, vars(vs, am), all_vars(. == 1))
```

All of the scoped filter functions have an option for `group_check`


### Mutating

The set of `reviewr` mutate functions report the number of NA or INF values created in the newly mutated variable(s).

```{r, echo = TRUE}
output <- mtcars %>% 
  mutate_qc(
    mpg_c21 = mpg - 21,
    inv_mpg_c21 = 1 / mpg_c21
  ) %>%
  select(mpg, mpg_c21, inv_mpg_c21)

head(output)
```

With grouped data and with `.group_check = T`, you can find out the number of missing rows per group and on which variables. With this option, a table will be printed with a columns called "var_name" and n_missing, indicating how many missing values there are for each newly mutated variable for each group. Note that this could print a big table if you have a lot of groups.

```{r, echo = TRUE}
result_normal <- mtcars %>% 
  group_by(vs) %>%
  mutate(
    mpg_c21 = mpg - 21,
    inv_mpg_c21 = 1 / mpg_c21,
    inv_am = 1 / am
  )

result_withOption <- mtcars %>% 
  group_by(vs) %>%
  mutate_qc(
    mpg_c21 = mpg - 21,
    inv_mpg_c21 = 1 / mpg_c21,
    inv_am = 1 / am,
    .group_check = T
  )
```

Whether or not `group_check = T`, the returned mutated data is identical. `.group_check = T` simply prints more information. The default is `.group_check = F`

```{r, echo = TRUE}
all.equal(result_normal, result_withOption)
```

`reviewr` has  `_qc` versions of the scoped filter functions, too.

```{r, echo = TRUE}
output <- mutate_all_qc(mtcars, funs(inv = 1 / .))
```

All of the scoped filter functions have an option for `.group_check`.

`reviewr` has `_qc` versions of `transmute()` and all of its scoped variants, as well.


### Summarizing

The set of `reviewr` summarize functions report the number of NA or INF values created in the newly summarized variable(s). Though available, this feature is typically not useful when using `summarize()` to create small summary tables - it will be pretty obvious in these cases when something is `NA` - but could be helpful when using `summarize()` on grouped data to create new variables you'll later use.

```{r, echo = TRUE}
mtcars %>% 
  group_by(vs) %>%
  summarize_qc(mean_inv_mpg_c21 = mean(1 / (mpg - 21)))
```

In this case, the `.group_check = T` option prints a table with each group with at least one missing value, and a single variable listing the newly summarized variables for the group that are missing. (Because `summarize()` by design only creates one row per group, there is no need to indicate the number of missing values, just flag which variables are missing.) Note that only groups with at least one missing value are printed.

```{r, echo = TRUE}
result <- mtcars %>% 
  group_by(vs) %>%
  summarize_qc(mean_inv_mpg_c21 = mean(1 / (mpg - 21)), .group_check = T)
```

Whether or not `group_check = T`, the returned summarized data is identical. `.group_check = T` simply prints more information. The default is `.group_check = F`.

`reviewr` has  `_qc` versions of the scoped filter functions, too.

```{r, echo = TRUE}
result <- mtcars %>%
  group_by(am) %>%
  summarize_at_qc(vars(mpg, hp, vs), funs(inv = mean(1 / .)))
```

All of the scoped filter functions have an option for `.group_check`.

`reviewr` has  `_qc` versions of `summarize()` and `summarise()` and all of their scoped variants. But this is America, so use a "z"!


### Joins

The set of `reviewr` join functions report several things:

* __Type of Join__  Joins are classified as one to one, one to many, many to one, or many to many. These definitions describe the extent to which there are duplicated rows of unique combinations of the `by` variable(s).

    * In _one to one_ merges, there is only one unique row of identifiers in each data set. Extra rows are never created in one to one joins.
    * _One to many_ and _many to one_ joins occur when one of the data sets has a duplicated id row. One to many implies the right data has the duplicated id; many to one implies the left data has the duplicated id. Additional rows may be created in these types of joins. 
    * _Many to many_ joins imply that both data sets have duplicated rows on the id variable(s). Additional rows may be created in this types of join.
    
* __Match Distribution in Joined Data__  The number of percent of rows in the newly joined data that were from the left only, right only, or matched.
* __Match Rates Based on Original Data__  The percent of the rows from the original left and right data frames that had a match.
* __Additional Rows__  When the join is not one-to-one, some rows might be duplicated. This indicates how many more rows than the original left and/or right data sets are in the newly joined data.

As with all the `reviewr` functions, simply add a `_qc` to the end of a `dplyr` join function:

```{r, echo = TRUE}
data_A <- 
  data.frame(
    id = 1:10, 
    var_A = 11:20
)
data_B <- 
  data.frame(
    id = c(5, 5, 5, 5, 6, 7, 7, 9, 10, 11),
    id_A = c(1:10),
    var_B = 21:30
  )

joined_data_normal <- left_join(data_A, data_B, by = "id")
joined_data_reviewr <- left_join_qc(data_A, data_B, by = "id")
```

If you do not use any of the additional options in the `reviewr` join functions (discussed below), the returned object is identical to `dplyr`.

```{r, echo = TRUE}
all.equal(joined_data_normal, joined_data_reviewr)
```

`reviewr` automatically detects the type of join:

```{r, echo = TRUE}
joined_data <- left_join_qc(data_A, data_B, by = c("id" = "id_A"))
```

There are two options available. `.merge` and `.extra`. Set `.merge` equal to a name you want to use for a new variable that will indicate, for each row, whether it was a match, from the left only, or from the right only.

```{r, echo = TRUE}
joined_data <- full_join_qc(data_A, data_B, .merge = "name_me")

print.data.frame(joined_data)
```

Set `.extra` equal to a name you want to use for a new variable that will indicate, for each row, whether it was a a combination of IDs that was duplicated from the left data, from the right data, or from both. These only occur in one to many, many to one, or many to many joins. In the example below, rows with the `id` of 5 from the left data was duplicated multiple times because 5 appeared on multiple rows in the right data. `,extra` flags these rows as "extra_on_left". The same happened with rows that had an `id` of 7.

```{r, echo = TRUE}
joined_data <- full_join_qc(data_A, data_B, .extra = "name_me_too")

print.data.frame(joined_data)
```

The defaults for both `.merge` and `.extra` are `NULL`. If either option is used, note that the returned object is no longer identical to `dplyr` because we've created new variables, though everything else beyond that is the same.

There are no group options for the `join` functions. As in `dplyr`, groups are ignored for the purposes of joining, but the groups from the first data set are preserved.


## What else?

I'd love to know what other types of quick QCs you find yourself doing repeatedly when manipulating data, especially those anchored to the `dplyr` functions. I expect and hope to widen this package going forward.